---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Ollama R Library

<!-- badges: start -->
[![R-CMD-check](https://github.com/hauselin/ollama-r/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/hauselin/ollama-r/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The [Ollama R library](https://hauselin.github.io/ollama-r/) provides the easiest way to integrate R with [Ollama](https://ollama.com/), which lets you run language models locally on your own machine. Main site: https://hauselin.github.io/ollama-r/

> Note: You should have at least 8 GB of RAM available to run the 7B models, 16 GB to run the 13B models, and 32 GB to run the 33B models.

See [Ollama's Github page](https://github.com/ollama/ollama) for more information. See also the [Ollama API documentation and endpoints](https://github.com/ollama/ollama/blob/main/docs/api.md). For Ollama Python, see [ollama-python](https://github.com/ollama/ollama-python). You'll need to have the [Ollama](https://ollama.com/) app installed on your computer to use this library. 

## Installation

1. You should have the Ollama app installed on your computer. Download it from [Ollama](https://ollama.com/).

2. Open/launch the Ollama app to start the local server. You can then run your language models locally, on your own machine/computer. 

3. Install the development version of `ollamar` R library like so:

``` r
devtools::install_github("hauselin/ollamar")
```

If it doesn't work or you don't have `devtools` installed, please run `install.packages("devtools")` in R or RStudio first. 

## Usage

`ollamar` uses the [`httr2` library](https://httr2.r-lib.org/index.html) to make HTTP requests to the Ollama server, so many functions in this library returns an `httr2_response` object by default. See [Notes section](#notes) below for more information.

```{r eval=FALSE}
library(ollamar)

test_connection()  # test connection to Ollama server; returns a httr2 response object
# Ollama local server running
# <httr2_response>

list_models()  # list available models (models you've pulled/downloaded)
   name                     size    parameter_size quantization_level modified           
   <chr>                    <chr>   <chr>          <chr>              <chr>              
 1 llama3:latest            4.7 GB  8B             Q4_0               2024-05-01T21:01:00
 2 mistral-openorca:latest  4.1 GB  7B             Q4_0               2024-04-25T16:45:00
```

### Pull/download model

Download a model from the ollama library (see [API doc](https://github.com/ollama/ollama/blob/main/docs/api.md#pull-a-model)). For the list of models you can pull/download, see [Ollama library](https://ollama.com/library).

```{r eval=FALSE}
pull("llama3")  # pull/download llama3 model
pull("mistral-openorca")  # pull/download mistral-openorca model
list_models()  # verify you've pulled/downloaded the model
```

### Delete a model

Delete a model and its data (see [API doc](https://github.com/ollama/ollama/blob/main/docs/api.md#delete-a-model)). You can see what models you've downloaded with `list_models()`. To download a model, specify the name of the model. 

```{r eval=FALSE}
list_models()  # see the models you've pulled/downloaded
delete("all-minilm:latest")  # returns a httr2 response object
```

### Generate a completion

Generate a response for a given prompt (see [API doc](https://github.com/ollama/ollama/blob/main/docs/api.md#generate-a-completion)).

```{r eval=FALSE}
resp <- generate("llama3", "Tomorrow is a...")  # return httr2 response object by default
resp

resp_process(resp, "text")  # process the response to return text/vector output

generate("llama3", "Tomorrow is a...", output = "text")  # directly return text/vector output
generate("llama3", "Tomorrow is a...", stream = TRUE)  # return httr2 response object and stream output
generate("llama3", "Tomorrow is a...", output = "df", stream = TRUE)
generate("llama3", "Tomorrow is a...", "text", TRUE)  # return text/vector output and stream output
```

### Chat

Generate the next message in a chat (see [API doc](https://github.com/ollama/ollama/blob/main/docs/api.md#generate-a-chat-completion)). See the [Notes section](#notes) for utility/helper functions to help you format/prepare the messages for the functions/API calls.

```{r eval=FALSE}
messages <- list(
    list(role = "user", content = "Who is the prime minister of the uk?")
)
resp <- chat("llama3", messages)  # default returns httr2 response object
resp  # <httr2_response>
chat("llama3", messages, "df")  # data frame/tibble
chat("llama3", messages, "raw")  # raw string
chat("llama3", messages, "jsonlist")  # list
chat("llama3", messages, "text")  # text vector

messages <- list(
    list(role = "user", content = "Hello!"),
    list(role = "assistant", content = "Hi! How are you?"),
    list(role = "user", content = "Who is the prime minister of the uk?"),
    list(role = "assistant", content = "Rishi Sunak"),
    list(role = "user", content = "List all the previous messages.")
)
chat("llama3", messages, "text")
```

#### Streaming responses

```{r eval=FALSE}
messages <- list(
    list(role = "user", content = "Hello!"),
    list(role = "assistant", content = "Hi! How are you?"),
    list(role = "user", content = "Who is the prime minister of the uk?"),
    list(role = "assistant", content = "Rishi Sunak"),
    list(role = "user", content = "List all the previous messages.")
)

# use "llama3" model, provide list of messages, return text/vector output, and stream the output
chat("llama3", messages, "text", TRUE)
# chat(model = "llama3", messages = messages, output = "text", stream = TRUE)  # same as above
```

### Embeddings

Get the vector embedding of some prompt/text (see [API doc](https://github.com/ollama/ollama/blob/main/docs/api.md#generate-embeddings)). By default, the embeddings are normalized to length 1, which means the following:

- cosine similarity can be computed slightly faster using just a dot product
- cosine similarity and Euclidean distance will result in the identical rankings

```{r eval=FALSE}
embeddings("llama3", "Hello, how are you?")

# don't normalize embeddings
embeddings("llama3", "Hello, how are you?", normalize = FALSE)
```

```{r eval=FALSE}
# get embeddings for similar prompts
e1 <- embeddings("llama3", "Hello, how are you?")
e2 <- embeddings("llama3", "Hi, how are you?")

# compute cosine similarity
sum(e1 * e2)  # 0.9859769
sum(e1 * e1)  # 1 (identical vectors/embeddings)

# non-normalized embeddings
e3 <- embeddings("llama3", "Hello, how are you?", normalize = FALSE)
e4 <- embeddings("llama3", "Hi, how are you?", normalize = FALSE)
sum(e3 * e4)  # 23695.96
sum(e3 * e3)  # 24067.32
```

### Notes

#### Optional/advanced parameters

Optional/advanced parameters (see [API docs](https://github.com/ollama/ollama/blob/main/docs/api.md)) such as `temperature` are not yet implemented as of now but will be added in the near future.

If you don't have the Ollama app running, you'll get an error. Make sure to open the Ollama app before using this library.

```{r eval=FALSE}
test_connection()
# Ollama local server not running or wrong server.
# Error in `httr2::req_perform()` at ollamar/R/test_connection.R:18:9:
```

#### Parsing `httr2_response` objects with `resp_process()`

`ollamar` uses the [`httr2` library](https://httr2.r-lib.org/index.html) to make HTTP requests to the Ollama server, so many functions in this library returns an `httr2_response` object by default. 

You can either parse the output with `resp_process()` or use the `output` parameter in the function to specify the output format. Generally, the `output` parameter can be one of `"df"`, `"jsonlist"`, `"raw"`, `"resp"`, or `"text"`.

```{r eval=FALSE}
resp <- list_models(output = "resp")  # returns a httr2 response object
# <httr2_response>
# GET http://127.0.0.1:11434/api/tags
# Status: 200 OK
# Content-Type: application/json
# Body: In memory (5401 bytes)

# process the httr2 response object with the resp_process() function
resp_process(resp, "df")
# or list_models(output = "df")
resp_process(resp, "jsonlist")  # list
# or list_models(output = "jsonlist")
resp_process(resp, "raw")  # raw string
# or list_models(output = "raw")
resp_process(resp, "resp")  # returns the input httr2 response object
# or list_models() or list_models("resp")
resp_process(resp, "text")  # text vector
# or list_models("text")
```

#### Utility/helper functions to format and prepare messages for the `chat()` function

Internally, messages are represented as a `list` of many distinct `list` messages. Each list/message object has two elements: `role` (either "user" or "assistant" or "system) and `content` (the message text). The example below shows how the messages/lists are presented. 

```{r eval=FALSE}
list(  # main list containing all the messages
    list(role = "user", content = "Hello!"),  # first message as a list
    list(role = "assistant", content = "Hi! How are you?"),  # second message as a list
    list(role = "user", content = "Who is the prime minister of the uk?"),  # third message as a list
    list(role = "assistant", content = "Rishi Sunak"),  # fourth message as a list
    list(role = "user", content = "List all the previous messages.")  # fifth message as a list
)
```

To simplify the process of creating and managing messages, `ollamar` provides utility/helper functions to format and prepare messages for the `chat()` function.

- `create_message()` creates the first message
- `append_message()` adds a new message to the end of the existing messages
- `prepend_message()` adds a new message to the beginning of the existing messages
- `insert_message()` inserts a new message at a specific index in the existing messages
    - by default, it inserts the message at the -1 (final) position
- `delete_message()` delete a message at a specific index in the existing messages
    - positive and negative indices/positions are supported
    - if there are 5 messages, the positions are 1 (-5), 2 (-4), 3 (-3), 4 (-2), 5 (-1)

```{r eval=FALSE}
# create first message
messages <- create_message(content = "Hi! How are you? (1ST MESSAGE)", role = "assistant")
# or simply, messages <- create_message("Hi! How are you?", "assistant")
messages[[1]]  # get 1st message

# append (add to the end) a new message to the existing messages
messages <- append_message("I'm good. How are you? (2ND MESSAGE)", "user", messages)
messages[[1]]  # get 1st message
messages[[2]]  # get 2nd message (newly added message)

# prepend (add to the beginning) a new message to the existing messages
messages <- prepend_message("I'm good. How are you? (0TH MESSAGE)", "user", messages)
messages[[1]]  # get 0th message (newly added message)
messages[[2]]  # get 1st message
messages[[3]]  # get 2nd message

# insert a new message at a specific index/position (2nd position in the example below)
# by default, the message is inserted at the end of the existing messages (position -1 is the end/default)
messages <- insert_message("I'm good. How are you? (BETWEEN 0 and 1 MESSAGE)", "user", messages, 2)
messages[[1]]  # get 0th message
messages[[2]]  # get between 0 and 1 message (newly added message)
messages[[3]]  # get 1st message
messages[[4]]  # get 2nd message

# delete a message at a specific index/position (2nd position in the example below)
messages <- delete_message(messages, 2)
```
