---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Ollama R Library

<!-- badges: start -->
<!-- badges: end -->

The Ollama R library provides the easiest way to integrate R with [Ollama](https://ollama.com/), which lets you run language models locally on your own machine. For Ollama Python, see [ollama-python](https://github.com/ollama/ollama-python). You'll need to have the [Ollama](https://ollama.com/) app installed on your computer to use this library. 

> Note: You should have at least 8 GB of RAM available to run the 7B models, 16 GB to run the 13B models, and 32 GB to run the 33B models.

See [Ollama's Github page](https://github.com/ollama/ollama) for more information. See also the [Ollama API documentation and endpoints](https://github.com/ollama/ollama/blob/main/docs/api.md).

## Installation

1. You should have the Ollama app installed on your computer. Download it from [Ollama](https://ollama.com/).

2. Open/launch the Ollama app to start the local server. You can then run your language models locally, on your own machine/computer. 

3. Install the development version of `ollamar` R library like so:

``` r
devtools::install_github("hauselin/ollamar")
```

If it doesn't work or you don't have `devtools` installed, please run `install.packages("devtools")` in R or RStudio first. 

## Usage

```{r eval=FALSE}
library(ollamar)

test_connection()  # test connection to Ollama server; returns a httr2 response object
# Ollama local server running
# <httr2_response>

list_models()  # list available models (models you've pulled/downloaded)
# A tibble: 16 × 4
   name                     model                    parameter_size quantization_level
   <chr>                    <chr>                    <chr>          <chr>             
 1 mixtral:latest           mixtral:latest           47B            Q4_0              
 2 llama3:latest            llama3:latest            8B             Q4_0              
```

### Notes

Optional/advanced parameters (see [API docs](https://github.com/ollama/ollama/blob/main/docs/api.md)) such as `temperature` are not yet implemented as of now but will be implemented in future versions.

If you don't have the Ollama app running, you'll get an error. Make sure to open the Ollama app before using this library.

```{r eval=FALSE}
test_connection()
# Ollama local server not running or wrong server.
# Error in `httr2::req_perform()` at ollamar/R/test_connection.R:18:9:
```

If a function in the library returns an `httr2_response` object, you can parse the output with `resp_process()`. `ollamar` uses the [`httr2` library](https://httr2.r-lib.org/index.html) to make HTTP requests to the Ollama server. 

```{r eval=FALSE}
resp <- list_models(output = "resp")  # returns a httr2 response object

# process the httr2 response object with the resp_process() function
resp_process(resp, "df")
resp_process(resp, "jsonlist")  # list
resp_process(resp, "raw")  # raw string
resp_process(resp, "resp")  # returns the input httr2 response object
```

### Pull/download model

Download a model from the ollama library (see [API doc](https://github.com/ollama/ollama/blob/main/docs/api.md#pull-a-model)). For the list of models you can pull/download, see [Ollama library](https://ollama.com/library).

```{r eval=FALSE}
pull("llama3")  # returns a httr2 response object
pull("mistral-openorca")
list_models("df")  # verify you've downloaded the model
```

### Delete a model

Delete a model and its data (see [API doc](https://github.com/ollama/ollama/blob/main/docs/api.md#delete-a-model)). You can see what models you've downloaded with `list_models()`. To download a model, specify the name of the model. 

```{r eval=FALSE}
list_models("df")  # see the models you've pulled/downloaded
delete("all-minilm:latest")  # returns a httr2 response object
```

### Chat

Generate the next message in a chat (see [API doc](https://github.com/ollama/ollama/blob/main/docs/api.md#generate-a-chat-completion)).

```{r eval=FALSE}
messages <- list(
    list(role = "user", content = "Who is the prime minister of the uk?")
)
chat("llama3", messages)  # returns a httr2 response object
chat("llama3", messages, output = "df")  # data frame/tibble
chat("llama3", messages, output = "raw")  # raw string
chat("llama3", messages, output = "jsonlist")  # list

messages <- list(
    list(role = "user", content = "Hello!"),
    list(role = "assistant", content = "Hi! How are you?"),
    list(role = "user", content = "Who is the prime minister of the uk?"),
    list(role = "assistant", content = "Rishi Sunak"),
    list(role = "user", content = "List all the previous messages.")
)
chat("llama3", messages)
```

#### Streaming responses

```{r eval=FALSE}
messages <- list(
    list(role = "user", content = "Hello!"),
    list(role = "assistant", content = "Hi! How are you?"),
    list(role = "user", content = "Who is the prime minister of the uk?"),
    list(role = "assistant", content = "Rishi Sunak"),
    list(role = "user", content = "List all the previous messages.")
)
chat("llama3", messages, stream = TRUE)
```

### Embeddings

Get the vector embedding of some prompt/text (see [API doc](https://github.com/ollama/ollama/blob/main/docs/api.md#generate-embeddings)). By default, the embeddings are normalized to length 1, which means the following:

- cosine similarity can be computed slightly faster using just a dot product
- cosine similarity and Euclidean distance will result in the identical rankings

```{r eval=FALSE}
embeddings("llama3", "Hello, how are you?")

# don't normalize embeddings
embeddings("llama3", "Hello, how are you?", normalize = FALSE)
```

```{r eval=FALSE}
# get embeddings for similar prompts
e1 <- embeddings("llama3", "Hello, how are you?")
e2 <- embeddings("llama3", "Hi, how are you?")

# compute cosine similarity
sum(e1 * e2)  # 0.9859769
sum(e1 * e1)  # 1 (identical vectors/embeddings)

# non-normalized embeddings
e3 <- embeddings("llama3", "Hello, how are you?", normalize = FALSE)
e4 <- embeddings("llama3", "Hi, how are you?", normalize = FALSE)
sum(e3 * e4)  # 23695.96
sum(e3 * e3)  # 24067.32
```

### Generate a completion

Generate a response for a given prompt (see [API doc](https://github.com/ollama/ollama/blob/main/docs/api.md#generate-a-completion)).

```{r eval=FALSE}
generate("llama3", "Tomorrow is a...", stream = TRUE)
generate("llama3", "Tomorrow is a...", output = "df")
generate("llama3", "Tomorrow is a...", stream = TRUE, output = "df")
```
